Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: The Case of the Palomar Transient Factory

### 2013-06-07 text/paper/ms.tex
v6

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged \apwsim40 times in \textit{R}-band, \apwsim$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations, a total of $1.1\times10^9$ light curves, to uncover three candidate sparse-field microlensing events. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large datasets, both of which will be useful to future wide-field, time-domain surveys such as that planned with the Large Synoptic Survey Telescope.

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005} and, most recently, to identify extrasolar planets \citep[][and references therein]{gaudi2011}. These studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year. In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects such as low-mass stars, sub-stellar objects, and isolated neutron stars.

%The microlensing event rate for a given region of the sky depends on the space and velocity distributions of the sources and on the space, velocity, and mass distribution of the lenses.
To maximize the event rate, microlensing surveys typically focus on high-density stellar regions such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.][]{original_ogle, original_macho, eros_original, crotts1996}. Of course, microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the event rate in low-density stellar regions will certainly be lower than that near e.g., the Bulge \citep[][]{wood_optical_depth, ogle_optical_depth, macho_optical_depth, eros_optical_depth}, these events are expected to involve closer sources and lenses and less crowded backgrounds compared to those found by modern microlensing surveys \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the event.

@@

To date, however, only one event outside of a dense stellar field has been recorded: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of a microlensing event of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Usually, such an analysis is impossible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and therefore mass measurements --- challenging. By contrast, high Galactic latitude events probe lenses with distances $\lesssim 1$~kpc, thus the Einstein radii tend to be larger (a few milliarcseconds) and may cause detectable astrometric signatures. A measurement of the astrometric shift of a source due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct lens mass measurement. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would therefore be very valuable.

In this paper, we use the Palomar Transient Factory (PTF) survey, described in Section 2, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. We examine the selection methods used by microlensing surveys to identify event signatures in light curves and an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~4 we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~5.

\section{PTF: Survey Design and Consequences for Microlensing Searches} 
PTF is an automated transient detection system \citep{nick2009,rau2009}. Data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48 inch Oschin Schmidt telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey $g$ \citep[SDSS;][]{york00} achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2010}. The data are processed in real time by a pipeline that identifies transients of interest and passes these to a dedicated photometric follow-up telescope; this has proved an excellent tool for e.g., early detection of supernovae \citep{nugent2011}. In addition, all detected sources are archived\footnote{The first photometric catalog of non-variable sources drawn from this archive is described in \citet[][]{ofek2012}.}; this enables the study of periodic forms of variability \citep[e.g., stellar or asteroid rotation;][]{agueros11,polishook2012}.

Images are processed using standard reduction procedures including de-biasing, flat-fielding, and astrometric calibration (Laher et al. in prep). Catalogs are generated using SExtractor for source identification \citep{bertin96}. The absolute photometric calibration of such images is good to a systematic limit of $\sim2\%$ for photometric nights, as described in \cite{ofek2012}. The final light curves are produced using relative photometric calibration which refines the calibration to $<1\%$ for photometric nights and improves the calibration for bad nights \citep[Levitan et al., in prep; for algorithm details see][]{ofek2012, levitan2011}.

As of 2012 Dec, the PTF footprint includes $\apwsim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$-band and $\apwsim$5430 (290) deg$^2$ imaged that often in $g$-band (see Figure~\ref{fig:survey_footprint})\footnote{For an interactive visualization of the PTF survey footprint, see \url{http://bit.ly/PTFSky}.}. The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by which of the PTF sub-surveys it belongs to, which time of the year it is visible from Palomar, and how high a priority it is given by the scheduler. The survey cadences mostly range from 1-5 days but include a high-cadence imaging campaign of a field in Orion to find transiting planets around young stars \citep{nick2009}. The result is that PTF light curves often contain gaps and regions of high-cadence observations and/or of low-cadence observations, so that the archive is a massive dataset of irregularly-sampled, time-domain photometry. Figure~\ref{fig:sampling} shows six randomly selected light curves and illustrates the varying cadences and coverage that different fields may have over the same one-year period.

@@

What would a microlensing event look like in a typical PTF light curve? If the source is an unblended point source and the lens is a foreground, dim object, a microlensing event is fully described by three parameters: the angular impact parameter $u_0$, the peak time of the event $t_0$, and the timescale of the event (Einstein crossing time) $t_E$. The amplification factor $A$ and flux $F$ can then be defined as a function of time: 
\begin{align}
	u(t) &= \sqrt{u_0^2 + 2\Big(\frac{t-t_0}{t_E}\Big)},\\
	A(t) &= \frac{u^2 + 2}{u\sqrt{u^2 + 4}},\\
	F(t) &= A(t)\times F_{source}.
\end{align}
The microlensing perturbation can also be expressed as:
\begin{align}\label{eq:ml_model}
	m(t) &= m_0(t) - 2.5\log A(t)
\end{align}
where $m_0(t)$ is the unperturbed, but possibly time-variable, magnitude of the source.

\begin{figure}[!t]
\begin{center}
\includegraphics[width=0.45\textwidth, trim=2 2 5 5, clip]{figures/sampling_figure.pdf}
\caption{Sampling of six different PTF fields over the course of a year. Darker points indicate that more exposures were taken that night; the total number of observations for each field is indicated.}\label{fig:sampling}
\end{center}
\end{figure}

\begin{figure}[h]
\centering\includegraphics[width=0.5\textwidth, trim=21 20 25 70, clip]{figures/num_observations_baseline_90deg.pdf}
\caption{Total number of $R$-band exposures and baselines (number of days between first and last exposure) for all PTF fields. The histogram to the right shows that the exposure distribution is peaked at 30-40; the histogram at the top shows that the baseline distribution has three peaks, corresponding to fields observed over $\lapprox$weeks, $\sim$1 year, and $\sim$3 years (the full length of the survey).}
\label{fig:num_obs}
\end{figure}

%However, if the event is \textit{too} bright, the pixels will saturate and there could be photometric errors flagging the most interesting data as bad.
Even in cases of high amplification ($u_0<<1$), a survey may miss or poorly sample an event if $t_E$ is short ($t_E \lesssim\mathrm{days}$), while if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. Figure~\ref{fig:microlensing_sim} illustrates the effect of simulated microlensing events with a fixed $t_E=20~\mathrm{days}$ but different $u_0$ on a random well-sampled PTF light curve.

\begin{figure}[!h]
\centering\includegraphics[width=0.48\textwidth, trim=5 5 5 7, clip]{figures/simulated_events.pdf}
\caption{Simulating the effect of a $20~\mathrm{day}$ microlensing event with differing impact parameters $u_0$ on a PTF light curve: a) the original light curve, b) the light curve with a $u_0=1.0$ event, c) with a $u_0=0.5$ event, d) with a $u_0=0.01$ event. The dashed line in the bottom panel shows the approximate saturation limit of the PTF camera in the $R$-band; such an event with a brighter source might therefore be missed by the survey.}\label{fig:microlensing_sim}
\end{figure}

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis \citep{alard1998} to identify transient events in raw imaging data. Light curves of any transient sources are then analyzed and vetted using a variety of selection methods to search for microlensing event candidates and distinguish them from e.g., variable stars, outbursting systems, and novae. Surveys have approached this process differently (as described in \citealt{ogle_optical_depth, con_idx, alcock2000, macho_detection_efficiency, udalski03,  hamadache2009, wyrzykowski2009, sumi2011}), but the general idea is to require that: 
\begin{enumerate}
	\item any selected light curve has some number of consecutive datapoints brighter than some threshold,
	\item a microlensing model best describe the data (compared to flat or linear light-curve models), as determined by using a $\Delta\chi^2$ test, and
	\item the model event parameters have physically reasonable values.
\end{enumerate}
%Difference imaging is ideal for studying crowded fields where photometric measurements are difficult. It also limits the number of light curves that must be fit with a model, thus saving on computation time and allowing for real-time detection of events, so that high-cadence follow-up observations of interesting sources can be triggered.

Applying this prescription to the PTF data presents obvious challenges. For example, microlensing surveys have relatively uniform time sampling of their survey footprint over an observing season, justifying the first requirement above. But it is harder to motivate such a cut on data with significant and irregular gaps. In order to develop the most successful procedure for PTF data, we therefore examine the relative performance of a set of variability indices and of the traditional prescription in selecting simulated microlensing events.

@@

Figure~\ref{fig:indices_examples} shows maximal outlier light curves for each variability statistic selected from \apwsim20,000 light curves for objects on a single CCD in PTF field 100101. %, centered at (J2000) 17:21:31.44 $+$61:07:46.6. 
Clearly, the indices are sensitive to different aspects of variability in the data. We expect $\sigma/\mu$ to be most useful for discerning periodic or semi-periodic variability where the variance is large; the expectations are not as clear for other indices.

\begin{figure}
\centering\includegraphics[width=.48\textwidth, trim=130 85 90 65, clip]{figures/max_outlier_light_curves.pdf}
\caption{The maximum outlier light curves for each variability index applied to data from PTF field 100101. For $\eta$ and $K$, this corresponds to the light curve with the \emph{minimum} value of the index over the entire field. For the other indices -- $\Delta\chi^2$, $J$, $\sigma/\mu$, $Con$ -- these are the light curves with the \emph{maximum} value of the relevant index over the field. } \label{fig:indices_examples}
\end{figure}

We conduct Monte Carlo simulations and inject artificial microlensing events into real PTF light curves. We then compare the distributions of variability indices for a set of light curves with and without these simulated events to determine the regions of parameter space where microlensing light curves fall. Figure~\ref{fig:var_indices} shows projections of the distribution for various combinations of the variability indices for another PTF field. These simulations do define regions to search for microlensing events, but there is still a large amount of overlap between the distributions with and without microlensing events. Below we describe our method for defining selection boundaries for each variability statistic to determine which is most efficient for detecting microlensing events.

% ^ discuss how the distributions change with injecting microlensing events

@@

\subsection{Event Selection and Detection Efficiency} \label{sec:detection_eff}
%The probability of detecting a microlensing event depends on the parameters of the event, the survey properties (i.e., observing strategy, limiting magnitude, weather), and the nature of the stellar field (i.e., amount of crowding). Before computing the detection efficiency, $\varepsilon$, for a given selection method,
We begin by randomly sampling 1000 light curves from each chip in a given PTF field. At least half of the individual observations in these light curves have to be defined as ``good'' \citep[see description of processing pipeline in][]{nick2009}. The light curves we reject are for objects that are either faint or near bright stars whose scattered light and diffraction spikes cause large photometric errors.

@@

The parameters for each simulated event are chosen as follows: $t_0$ is drawn from a uniform distribution between the minimum and maximum observation times, $u_0$ is drawn from a uniform distribution between 0 and \apwsim1.5 \citep[the impact parameter that causes a maximum deviation larger than \apwsim$5\sigma$ for an $R\sim17.5$ mag source; e.g.,][]{macho_detection_efficiency}, and $t_E$ is drawn from a log-uniform distribution between 1 and 1000 days. For each iteration we recompute the variability indices and store the event parameters.

Figures~\ref{fig:detection_efficiency_4327}-\ref{fig:detection_efficiency_100152} show the results for three representative fields with different sampling patterns (top panels) and detection efficiency curves computed using this simulation (bottom panels). We ignore the indices \textit{K}, $\sigma/\mu$ and $Con$ because their integrated detection efficiencies are below 1\%. At a fixed FPR of 1\%, $Con$ performs poorly, but it may be useful as an initial cut by allowing a higher FPR. In this work, we only consider a single-index selection method and thus reject $Con$. For each of these fields we find that $\eta$ consistently performs better than the other indices at recovering microlensing events. This is especially interesting because computing $\eta$ is $\apwsim100\times$ faster per light curve than computing $\Delta\chi^2$, and that identifying candidate microlensing events in the full PTF dataset using this statistic is therefore computationally plausible.

@@

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {figures/field004327_Nperccd1000_Nevents100.pdf}
	   \label{fig:subfig2}
	 }
\caption{{\it Top:} A randomly selected light curve from PTF field 4327. Note the sampling pattern. {\it Bottom:} Detection efficiency $\varepsilon$ for $\eta$, $\Delta\chi^2$, and $J$ as functions of the simulated microlensing event parameters $t_E$, $u_0$, and $m_0$, the microlensing event timescale, angular impact parameter, and `quiescent' source magnitude, respectively. Black (dashed) lines show the normalized input distributions, red (solid) lines show the recovered distributions.}  \label{fig:detection_efficiency_4327}
\end{figure}

@@

\section{Searching for events in the full PTF dataset}
To identify candidate microlensing events in the $1.1\times10^9$ light curves with $>$10 $R$-band observations, we:
%We apply the 1\% FPR cut to the 2097 PTF fields (\apwsim15,224 deg$^2$) with $>$10 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select a field from the 2097 fields with $>$10 good $R$-band observations and run Monte Carlo simulations to compute the 1\% FPR limiting values for $\eta$, $\Delta \chi^2$, and $J$ for that field;
	\item compute $\eta$ for all the light curves in the field, and select the light curves that pass the quality cut described in Section~\ref{sec:detection_eff} and the FPR cut determined from the Monte Carlo simulations;
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for all the selected light curves;
	\item reject the light curves for which the new value of $\eta$ still passes the cut\footnote{These are probably periodic variables.};
	\item retain any remaining light curves for further inspection.
\end{enumerate}
% Turn this into a table, with number of objects left over after each cut?

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we use Data Release 8 photometry \citep{DR8paper} to identify candidate quasars \citep{richards02} and examine the SDSS images to remove extended objects \citep{sdss_images}. Removing extragalactic objects from our candidate list reduces contamination by e.g., supernovae.

We visually inspect each of the remaining \apwsim2000 light curves. We identify \apwsim1100 quasars that did not meet the \cite{richards02} color cut, or objects with unknown long-term variability (e.g., Mira-type variables). An additional \apwsim600 light curves have bad or poorly calibrated data that mimics a transient increase in flux: diffraction spikes, ghosts, and scattered light can cause non-Gaussian, seeing-dependent variability. Where appropriate, as an additional test, we examine the PTF images to verify the data quality. A small fraction of the selected light curves are variable stars that made it past our periodic variability cut.

>>>>

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, performs best for identifying simulated microlensing events. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations, a total of $1.1\times10^9$ light curves, to uncover three candidate sparse-field microlensing events. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large datasets, both of which will be useful to future wide-field, time-domain surveys such as that planned with the Large Synoptic Survey Telescope.

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and, most recently, to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year. In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects such as low-mass stars, sub-stellar objects, and isolated neutron stars.

%The microlensing event rate for a given region of the sky depends on the space and velocity distributions of the sources and on the space, velocity, and mass distribution of the lenses.
To maximize the event rate, microlensing surveys typically focus on high-density stellar regions such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.][]{original_ogle, original_macho, eros_original, crotts1996}. Of course, microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the event rate in low-density stellar regions will certainly be lower than that near e.g., the Bulge \citep[][]{ogle_optical_depth, macho_optical_depth, wood_optical_depth, eros_optical_depth}, these events are expected to involve closer sources and lenses, and less crowded backgrounds, compared to those found by modern microlensing surveys \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and therefore mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii therefore tend to be larger (a few milliarcseconds) and may cause detectable astrometric signatures. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would therefore be very valuable.

In this paper, we use the Palomar Transient Factory (PTF) survey, described in Section~\ref{sec:ptf}, as a test case for studying the frequency and detectability of microlensing events in all-sky, irregularly sampled, time-domain surveys. We examine the selection methods used by microlensing surveys to identify event signatures in light curves and an event identification procedure based on a set of variability statistics. We then present the detection efficiency for these methods as computed for a representative set of PTF fields (Section~\ref{sec:event_recovery}). In Section~\ref{sec:search} we apply a set of criteria that use the results of these tests to all PTF light curves with $>$10 high-quality observations to identify candidate microlensing events. We discuss eight interesting transients and three candidate events identified in this manner before concluding in Section~\ref{sec:conc}.

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF is an automated transient detection system \citep{nick2009,rau2009}. Data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48 inch Oschin Schmidt telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (1.1$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2010}. The data are processed in real time by a pipeline that identifies transients of interest and passes these to a dedicated photometric follow-up telescope; this has proved an excellent tool for e.g., early detection of supernovae \citep{nugent2011}.

Images are processed using standard reduction procedures including de-biasing, flat-fielding, and astrometric calibration (Laher et al., in prep). Catalogs are then generated using SExtractor for source identification \citep{bertin96}. The absolute photometric calibration of the images is good to a systematic limit of $\sim2\%$ for photometric nights, as described in \cite{ofek2012}.  The final light curves are produced using relative photometric calibration, which refines the calibration to $<$1\% for photometric nights and improves the calibration for bad nights \citep[Levitan et al., in prep; for algorithm details see][]{levitan2011, ofek2012}. The resulting archive is an excellent resource for the study of periodic forms of variability \citep[e.g., stellar or asteroid rotation;][]{agueros11,polishook2012}.

%In addition, all detected sources are archived\footnote{The first photometric catalog of non-variable sources drawn from this archive is described in \citet[][]{ofek2012}.};

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.47\textwidth, trim=2 2 5 5, clip]{figures/sampling_figure.pdf}
\caption{Sampling of six different PTF fields over the course of a year. Darker points indicate that more exposures were taken that night; the total number of observations for each field is indicated.}\label{fig:sampling}
\end{center}
\end{figure}

\begin{figure}[t]
\centering\includegraphics[width=0.5\textwidth, trim=21 20 25 70, clip]{figures/num_observations_baseline_90deg.pdf}
\caption{Total number of $R$-band exposures and baselines (number of days between first and last exposure) for all PTF fields. The histogram to the right shows that the exposure distribution is peaked at 30-40; the histogram at the top shows that the baseline distribution has three peaks, corresponding to fields observed over $\lapprox$weeks, $\sim$1 year, and $\sim$3 years (the full length of the survey).}
\label{fig:num_obs}
\end{figure}

\begin{figure}[!h]
\centering\includegraphics[width=0.48\textwidth, trim=5 5 5 7, clip]{figures/simulated_events.pdf}
\caption{Simulating the effect of $20$-day microlensing events with impact parameters $u_0 = 1.0$, 0.5, and 0.01, on a random PTF $R$-band light curve. The original light curve is shown at the top. The dashed line in the bottom panel shows the approximate saturation limit of the PTF camera in the $R$-band; such an event with a brighter source might therefore be missed by the survey.}\label{fig:microlensing_sim}
\end{figure}

As of 2012 Dec, the PTF footprint includes $\sim$15,224 (2766) deg$^2$ imaged $>$10 ($>$100) times in $R$-band and $\sim$5430 (290) deg$^2$ imaged that often in $g$-band (see Figure~\ref{fig:survey_footprint})\footnote{For an interactive visualization of the PTF survey footprint, see \url{http://bit.ly/PTFSky}.}. The PTF survey footprint is not uniformly sampled either spatially or temporally. Each field has a unique sampling pattern determined by which of the PTF sub-surveys it belongs to, what time of the year it is visible, and how high a priority it is given by the scheduler. The imaging cadences range from 1-5 days, but there is a higher-cadence campaign of a field in Orion to find transiting planets around young stars \citep{nick2009}. The result is that PTF light curves often contain gaps and regions of high-cadence observations and/or of low-cadence observations, so that the archive is a massive dataset of irregularly-sampled, time-domain photometry. Figure~\ref{fig:sampling} shows six randomly selected light curves and illustrates the varying cadences and coverage that different fields may have over the same one-year period.

@@

What would a microlensing event look like in a typical PTF light curve? If the source is an unblended point source and the lens is a foreground, dim object, a microlensing event is fully described by three parameters: the angular impact parameter $u_0$, the peak time of the event $t_0$, and the timescale of the event (Einstein crossing time) $t_E$. The amplification factor $A$ and flux $F$ can then be defined as a function of time: 
\begin{align}
	u(t) &= \sqrt{u_0^2 + 2\Big(\frac{t-t_0}{t_E}\Big)},\\
	A(t) &= \frac{u^2 + 2}{u\sqrt{u^2 + 4}},\\
	F(t) &= A(t)\times F_{source}.
\end{align}
The microlensing perturbation can also be expressed as:
\begin{align}\label{eq:ml_model}
	m(t) &= m_0(t) - 2.5\log A(t),
\end{align}
where $m_0(t)$ is the unperturbed, but possibly time-variable, magnitude of the source.

%However, if the event is \textit{too} bright, the pixels will saturate and there could be photometric errors flagging the most interesting data as bad.
Even in cases of high amplification ($u_0<<1$), a survey may miss or poorly sample an event if $t_E$ is short ($t_E \lesssim\mathrm{days}$), while if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. Figure~\ref{fig:microlensing_sim} illustrates the effect of simulated microlensing events with a fixed $t_E=20~\mathrm{days}$, but different $u_0$, on a random well-sampled PTF light curve.

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis \citep{alard1998} to identify transient events in raw imaging data. Light curves of any transient sources are then analyzed and vetted using a variety of selection methods to search for microlensing event candidates and distinguish them from e.g., variable stars, outbursting systems, and novae. Surveys have approached this process differently (as described in \citealt{ogle_optical_depth, con_idx, alcock2000, macho_detection_efficiency, udalski03,  hamadache2009, wyrzykowski2009, sumi2011}), but the general idea is to require that: 
\begin{enumerate}
	\item any selected light curve has some number of consecutive datapoints brighter than some threshold,
	\item compared to flat or linear light-curve models, a microlensing model best describe the data, as determined by using a $\Delta\chi^2$ test, and
	\item the microlensing model event parameters have physically reasonable values.
\end{enumerate}
%Difference imaging is ideal for studying crowded fields where photometric measurements are difficult. It also limits the number of light curves that must be fit with a model, thus saving on computation time and allowing for real-time detection of events, so that high-cadence follow-up observations of interesting sources can be triggered.

Applying this prescription to the PTF data presents obvious challenges. For example, microlensing surveys have relatively uniform time sampling of their survey footprint over an observing season, justifying the first requirement. But it is harder to motivate such a cut on data with significant and irregular gaps. In order to develop the most successful procedure for PTF data, we therefore examine the relative performance of a set of variability indices and of the traditional prescription in selecting simulated microlensing events.

@@

Figure~\ref{fig:indices_examples} shows maximally outlying light curves for each variability statistic selected from $\sim$20,000 light curves for objects on a single CCD in PTF field 100101. %, centered at (J2000) 17:21:31.44 $+$61:07:46.6. 
Clearly, the indices are sensitive to different aspects of variability in the data. We expect $\sigma/\mu$ to be most useful for discerning periodic or semi-periodic variability where the variance is large; the expectations are not as clear for other indices.

\begin{figure}
\centering\includegraphics[width=.48\textwidth, trim=130 85 90 65, clip]{figures/max_outlier_light_curves.pdf}
\caption{The maximally outlying light curves for each variability index applied to data from PTF field 100101. For $\eta$ and $K$, this corresponds to the light curve with the \emph{minimum} value of the index over the entire field. For the other indices, these are the light curves with the \emph{maximum} value of the relevant index over the field. } \label{fig:indices_examples}
\end{figure}

We conduct Monte Carlo simulations and inject artificial microlensing events into real PTF light curves. We then compare the distributions of variability indices for a set of light curves with and without these simulated events to determine the regions of parameter space where microlensed light curves fall. Figure~\ref{fig:var_indices} shows projections of the distribution for various combinations of the variability indices for another PTF field. These simulations do define regions to search for microlensing events, but there is still a large amount of overlap between the distributions with and without microlensing events. Below we describe our method for defining selection boundaries for each variability statistic to determine which is most efficient for detecting microlensing events.

@@

\subsection{Event Selection and Detection Efficiency} \label{sec:detection_eff}
%The probability of detecting a microlensing event depends on the parameters of the event, the survey properties (i.e., observing strategy, limiting magnitude, weather), and the nature of the stellar field (i.e., amount of crowding). Before computing the detection efficiency, $\varepsilon$, for a given selection method,
We begin by randomly sampling 1000 light curves from each chip in a given PTF field. At least half of the individual observations in these light curves must be defined as ``good'' \citep[see description of processing pipeline in][]{nick2009}. The rejected light curves are for objects that are either faint or near bright stars whose scattered light and diffraction spikes cause large photometric errors.

@@

The parameters for each simulated event are chosen as follows: $t_0$ is drawn from a uniform distribution between the first and last observation date, $u_0$ is drawn from a uniform distribution between 0 and $\sim$1.5 \citep[the impact parameter that causes a maximum deviation larger than $\sim$$5\sigma$ for a $R\sim17.5$ mag source; e.g.,][]{macho_detection_efficiency}, and $t_E$ is drawn from a log-uniform distribution between 1 and 1000 days. For each iteration we recompute the variability indices and store the event parameters.

Figures~\ref{fig:detection_efficiency_4327}-\ref{fig:detection_efficiency_100152} show the results for three representative fields with different sampling patterns (top panels) and detection efficiency curves computed using this simulation (bottom panels). We ignore the indices \textit{K} and $\sigma/\mu$ because their integrated detection efficiencies are below 1\%. At a fixed FPR of 1\%, $Con$ performs poorly, but it may be useful as an initial cut if a higher FPR is used. In this work, we only consider a single-index selection method and thus reject $Con$.

For each of these fields we find that $\eta$ consistently performs better than the other indices at recovering microlensing events. This is especially interesting because computing $\eta$ is $\sim$100$\times$ faster per light curve than computing $\Delta\chi^2$. Identifying candidate microlensing events in the full PTF dataset using this statistic is therefore computationally plausible.

@@

	 \subfigure{
	   \includegraphics[width=.46\textwidth, trim=8 8 5 5, clip] {figures/field004327_Nperccd1000_Nevents100.pdf}
	   \label{fig:subfig2}
	 }
\caption{{\it Top:} A randomly selected light curve from PTF field 4327. Note the sampling pattern. {\it Bottom:} Detection efficiency $\varepsilon$ for $\eta$, $\Delta\chi^2$, and $J$, as a function of the simulated microlensing event parameters $t_E$, $u_0$, and $m_0$ (the event timescale, impact parameter, and `quiescent' source magnitude, respectively). Black (dashed) lines show the normalized input distributions, red (solid) lines show the recovered distributions.}  \label{fig:detection_efficiency_4327}
\end{figure}

@@

\section{Searching for events in the full PTF dataset}\label{sec:search}
To identify candidate microlensing events in the $1.1\times10^9$ light curves with $>$10 $R$-band observations, we:
%We apply the 1\% FPR cut to the 2097 PTF fields ($\sim$15,224 deg$^2$) with $>$10 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select a field from the 2097 fields with $>$10 good $R$-band observations and run Monte Carlo simulations to compute the 1\% FPR limiting values for $\eta$, $\Delta \chi^2$, and $J$ for that field;
	\item compute $\eta$ for all the light curves in the field, and select the light curves that pass the quality cut described in Section~\ref{sec:detection_eff} and the FPR cut determined from the Monte Carlo simulations;
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for all the selected light curves;
	\item reject the light curves for which the new value of $\eta$ still passes the cut\footnote{These are probably periodic variables.};
	\item retain any remaining light curves for further inspection.
\end{enumerate}
% Turn this into a table, with number of objects left over after each cut?

This procedure identifies 2377 candidates from among the initial sample of $1.1\times10^9$ light curves. We search SIMBAD\footnote{This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France.} and remove any candidate among these with a known extragalactic counterpart within 10\asec. For those in the SDSS footprint, we use Data Release 8 photometry \citep{DR8paper} to identify candidate quasars \citep[using the cuts described in][]{richards02} and examine the SDSS images \citep{sdss_images} to remove extended objects. Removing extragalactic objects in this manner from our candidate list reduces contamination by e.g., supernovae.

We visually inspect each of the remaining $\sim$2000 light curves. We identify $\sim$1100 quasars that did not meet the \cite{richards02} color cut, or objects with unknown long-term variability (e.g., Mira-type variables). An additional $\sim$600 light curves have bad or poorly calibrated data that mimics a transient increase in flux: diffraction spikes, ghosts, and scattered light can cause non-Gaussian, seeing-dependent variability. Where appropriate, as an additional test, we examine the PTF images to verify the data quality. A small fraction of the selected light curves are variable stars that made it past our periodic variability cut.
